---
title: "Lab 02 - Plastic waste"
author: "Olivia Zhang"
date: "1/27/2025"
output: github_document
---

## Load packages and data

```{r load-packages, message=FALSE}
library(tidyverse) 
```

```{r load-data, message=FALSE}
plastic_waste <- read.csv("data/plastic-waste.csv")
```

## Exercises

Examine the data.

```{r distribution plastic_waste_per_cap}
ggplot(data = plastic_waste, aes(x = plastic_waste_per_cap)) +
  geom_histogram(binwidth = 0.2)
```

Identify the outlier country.

```{r}
plastic_waste %>%
  filter(plastic_waste_per_cap > 3.5)
```


### Exercise 1

Distributions of plastic waste per capita for different continents.

```{r plastic-waste-continent}
ggplot(data = plastic_waste, aes(x = plastic_waste_per_cap)) +
  geom_histogram(binwidth = 0.2) +
  facet_wrap(~continent, ncol = 3)
```

```{r density-plot}
ggplot(
  data = plastic_waste,
  aes(x = plastic_waste_per_cap)
) +
  geom_density()
```

```{r density-plot-continent}
ggplot(
  data = plastic_waste,
  mapping = aes(
    x = plastic_waste_per_cap,
    color = continent
  )
) +
  geom_density()
```

```{r density-plot-continent-fill}
ggplot(
  data = plastic_waste,
  mapping = aes(
    x = plastic_waste_per_cap,
    color = continent,
    fill = continent
  )
) +
  geom_density()
```

### Exercise 2

```{r plastic-waste-density}
ggplot(
  data = plastic_waste,
  mapping = aes(
    x = plastic_waste_per_cap,
    color = continent,
    fill = continent
  )
) +
  geom_density(alpha = 0.2)
```

We define the color and fill of the curves by mapping aesthetics of the plot because it would give us legends regarding which color/fill represents which group, while there would be no legend for characteristics that are mapped in the plotting geom, like in the case of the alpha level.

### Exercise 3

```{r side-by-side box-plots}
ggplot(
  data = plastic_waste,
  mapping = aes(
    x = continent,
    y = plastic_waste_per_cap
  )
) +
  geom_boxplot()
```

Convert your side-by-side box plots from the previous task to violin plots. What do the violin plots reveal that box plots do not? What features are apparent in the box plots but not in the violin plots?

```{r plastic-waste-violin}
# insert code here
```

### Exercise 4

Remove this text, and add your answer for Exercise 4 here.

```{r plastic-waste-mismanaged}
# insert code here
```

```{r plastic-waste-mismanaged-continent}
# insert code here
```

```{r plastic-waste-population-total}
# insert code here
```

```{r plastic-waste-population-coastal}
# insert code here
```

### Exercise 5

Remove this text, and add your answer for Exercise 5 here.

```{r recreate-viz}
# insert code here
```


