---
title: "Lab 02 - Plastic waste"
author: "Olivia Zhang"
date: "1/27/2025"
output: github_document
---

## Load packages and data

```{r load-packages, message=FALSE}
library(tidyverse) 
```

```{r load-data, message=FALSE}
plastic_waste <- read.csv("data/plastic-waste.csv")
```

## Exercises

Examine the data.

```{r distribution plastic_waste_per_cap}
ggplot(data = plastic_waste, aes(x = plastic_waste_per_cap)) +
  geom_histogram(binwidth = 0.2)
```

Identify the outlier country.

```{r}
plastic_waste %>%
  filter(plastic_waste_per_cap > 3.5)
```


### Exercise 1

Distributions of plastic waste per capita for different continents.

```{r plastic-waste-continent}
ggplot(data = plastic_waste, aes(x = plastic_waste_per_cap)) +
  geom_histogram(binwidth = 0.2) +
  facet_wrap(~continent, ncol = 3)
```

```{r density-plot}
ggplot(
  data = plastic_waste,
  aes(x = plastic_waste_per_cap)
) +
  geom_density()
```

```{r density-plot-continent}
ggplot(
  data = plastic_waste,
  mapping = aes(
    x = plastic_waste_per_cap,
    color = continent
  )
) +
  geom_density()
```

```{r density-plot-continent-fill}
ggplot(
  data = plastic_waste,
  mapping = aes(
    x = plastic_waste_per_cap,
    color = continent,
    fill = continent
  )
) +
  geom_density()
```

### Exercise 2

```{r plastic-waste-density}
ggplot(
  data = plastic_waste,
  mapping = aes(
    x = plastic_waste_per_cap,
    color = continent,
    fill = continent
  )
) +
  geom_density(alpha = 0.2)
```

We define the color and fill of the curves by mapping aesthetics of the plot because it would give us legends regarding which color/fill represents which group, while there would be no legend for characteristics that are mapped in the plotting geom, like in the case of the alpha level.

### Exercise 3

```{r side-by-side box-plots}
ggplot(
  data = plastic_waste,
  mapping = aes(
    x = continent,
    y = plastic_waste_per_cap
  )
) +
  geom_boxplot()
```

Now convert the side-by-side box plots from the previous task to violin plots. 

```{r plastic-waste-violin}
ggplot(
  data = plastic_waste,
  mapping = aes(
    x = continent,
    y = plastic_waste_per_cap
  )
) +
  geom_violin()
```

The violin plots show the distributions of plastic waste per capita for different continents, while the side-by-side box plots clearly tells us about the means, quartiles, and outliers of plastic waste for different continents, allowing straightforward comparisons of descriptive statistics. 

### Exercise 4

The relationship between plastic waste per capita and mismanaged plastic waste per capita using a scatterplot is visualized below:

```{r plastic-waste-mismanaged}
ggplot(
  data = plastic_waste,
  mapping = aes(
    x = plastic_waste_per_cap,
    y = mismanaged_plastic_waste_per_cap
  )
) +
  geom_point()
```

As you can see here, the more the plastic waste per capita, the more mismanaged plastic waste per capita. The trend going up is pretty steep, which means that the positive correlation between the plastic waste per capita and the more mismanaged plastic waste per capita is pretty strong. However, there are also some variations of the trend, which could represent data coming from different continents where the strategies solving plastic waste is different, some more effective and others not. And of course, there is one outlier.

```{r plastic-waste-mismanaged-continent}
ggplot(
  data = plastic_waste,
  mapping = aes(
    x = plastic_waste_per_cap,
    y = mismanaged_plastic_waste_per_cap, 
    color = continent
  )
) +
  geom_point()
```

The color scatterplot reveals more information about how plastic waste per capita and mismanaged plastic waste per capita are associated with respect different continents. Africa seems to be the most inefficient plastic waster handler. Europe and some Asian countries seems to be very efficient in handling plastic waster, while others, North America, and South America are somewhere in the middle.

```{r plastic-waste-population-total}
ggplot(
  data = plastic_waste,
  mapping = aes(
    x = total_pop,
    y = plastic_waste_per_cap, 
    color = continent
  )
) +
  geom_point()
```

```{r plastic-waste-population-coastal}
ggplot(
  data = plastic_waste,
  mapping = aes(
    x = coastal_pop,
    y = plastic_waste_per_cap, 
    color = continent
  )
) +
  geom_point()
```

plastic waste per capita seems to be a little bit more linear related to the coastal population than the total population, but only by a function of a few Asian countries.

### Exercise 5

The relationship between the proportion of coastal population and plastic waste per capita:

```{r recreate-viz}
plastic_waste$coastal_pop_por = plastic_waste$coastal_pop / plastic_waste$total_pop

ggplot(plastic_waste %>% filter(plastic_waste_per_cap < 3),
  mapping = aes(
    x = coastal_pop_por,
    y = plastic_waste_per_cap, 
    color = continent
  )
) +
  geom_point()+
  labs(title = "Plastic waste vs. coastal population proportion", 
       subtitle = "by continent", 
       x = "Coastal population proportion (Coastal / total population)", 
       y = "Plastic waster per capita") +
  scale_color_viridis_d() +
  geom_smooth(color = 1)
```


